The Social Conditioning of Fear

### 2023-12-05 scripts/manuscript.qmd
Methods write-up

---
title: "Social Conditions of Fear"
author:
  - Alejandra Pedraza
  - W. Jamie Yang
abstract: "TBW" # this is where abstract should be pasted 
format: html
toc: true
toc-location: left
toc-title: Table of Contents
highlight-style: pygments
editor: visual
execute:
  eval: true # change to false before final submission
  echo: true
  message: false
  warning: false
fig-width: 6.5
fig-height: 8
code-fold: true
code-summary: "View Code"
embed-resources: true
bibliography: references.bib
link-citations: true
---

@@

This study draws data from the General Social Survey (GSS). The GSS is a nationally representative interview-based survey that has collected longitudinal data since 1972. The GSS contains questions on demographics, behaviors, attitudes, civil liberaties, crime and violence, intergroup tolerance, morality, national spending priorities, psychological well-being, social mobility, stress, and traumatic events. Our study is limited to data collected during the 2022 wave and we draw primarily on demographic and well-being variables.

# Appendices

## Appendix 1: Codebook

### Dependent Variable

-   **fear**: Is there any area right around here--that is, within a mile--where you would be afraid to walk alone at night?

### Independent Variables

#### Individual-Level

@@

-   **sexnow1**: Do you describe yourself as male, female, or transgender? *This is self-described gender.*

-   **sex**: Code respondent's sex (*This is gender presentation as interpreted by the interviewer).*

@@

-   **racerank1**: If you had to choose, which of these races do you identify most with? *Kept this race variable here instead of "race" because it allows more heterogeneity within group.*

@@

#### Community-Level

@@

## Appendix II: Data Cleaning

@@

library(tidyverse)
library(dplyr)
library(tidyr)
library(here)
library(rio)
library(janitor)
library(knitr)
library(kableExtra)
library(mice)
library(forcats)
library(ggplot2)
library(RColorBrewer)
```

@@

The GSS codes missing data as either ".i: Inapplicable", ".n: No answer", ".d: Do not Know/Cannot Choose", and ".s: Skipped on Web. For the purposes of this project, we are considering all of the above as missing values. For our next steps, we need to reexamine this decision as".i: Inapplicable" is a valid skip and needs to be accounted for accordingly.

In the following code chunk, we recode the aforementioned indicators as NAs. This is a necessary first step to run mice imputations.

@@

First we filtered out all NAs from our dependent variable, fear.

Informed by theory, race/ ethnicity is an important covariate for our models. First, we sought to create a combined race variable from the variables "race" and "hispanic." After looking at the data, we saw that the options for the race variable were "Black, White, Other." This does not provide the control we need. So, we opted to use the "racerank1" variable, because participants were able to select from 16 different options: White, Black or African-American, Asian Indian, Chinese, Filipino, Japanese, Korean, Vietnamese, Other Asian, Guamanian or Chamorro, Samoan, Other Pacific Islander, Some other race, and Hispanic. We factored and recoded these to White, Black, Hispanic, Asian, NHPI (Native Hawaiian or Pacific Islander), AIAN (American Indian, Alaska Native), and Other. We understand that these categories are problematic and contentious, however they are the standard with quantitative data.

Then we factored and recoded selected variables for our analysis. It was important to not have any variables as character strings, otherwise mice imputations would not work.

@@

Once our data was tidy, we selected the variables we would be using for our analysis.

@@

To deal with the large amount of missing values, we ran mice imputations. In the code below we create a new dataset with the imputed values. We will use this dataset in our next stage to run various models.

>>>>

---
title: "The Social Conditioning of Fear"
author:
  - Alejandra Pedraza
  - W. Jamie Yang
abstract: "TBW" # this is where abstract should be pasted 
format: html
toc: true
toc-location: left
toc-title: Table of Contents
highlight-style: pygments
editor: visual
execute:
  eval: true # change to false before final submission
  echo: true
  message: false
  warning: false
fig-width: 6.5
fig-height: 8
code-fold: true
code-summary: "View Code"
embed-resources: true
bibliography: references.bib
link-citations: true
---

@@

This study draws data from the General Social Survey (GSS). The GSS is a nationally representative interview-based survey that has collected longitudinal data since 1972. The GSS contains questions on demographics, behaviors, attitudes, civil liberties, crime and violence, intergroup tolerance, morality, national spending priorities, psychological well-being, social mobility, stress, and traumatic events. Our study is limited to data collected during the 2022 wave and we draw primarily on demographic and well-being variables.

Our dependent variable is coded as **fear** by the GSS. Respondents were asked, "*Is there any area right around here--that is, within a mile--where you would be afraid to walk alone at night*?" During the 2022 wave, there were 3,544 answers coded as follows: 878 = yes, 1440 = no, and 1226 = NA. Notably, the GSS codes missing values with multiple indicators--".i: Inapplicable," ".n: No answer," ".d: Do not Know/Cannot Choose," and ".s: Skipped on Web." For the purposes of our preliminary study, we coded all missing data as NA. Furthermore, we applied fixed removal of NAs for our dependent variable and we ran mice imputations for all other variables to create a more robust dataset that is necessary for further modeling.

Our control and independent variables are representative of two different dimensions we sought to capture. The first dimension is individual-level or identity characteristics:

@@

-   **sexnow1**: Do you describe yourself as male, female, or transgender?

-   **sex**: Code respondent's sex

@@

-   **racerank1**: If you had to choose, which of these races do you identify most with?

@@

Informed by theory, race is an important co-variate for our models as racialization serves to structure the daily lives of individuals based on where they fall on the Black-white spectrum, their opportunities, and their sense of belonging [@bonilla-silva2022; @omi2015]. First, we sought to create a combined race variable from the variables **race** and **hispanic**. After looking at the data, we saw that **race** was coded as "Black, White, or Other." This does not provide the control we need as it omits those in the "racial middle" that are more likely to identify with a panethnic label and reject the Black-white binary \[ @mora2014; @obrien2008; [@rodriguez2000]\]. Instead, we opted to use the **racerank1** variable, because participants were able to select from 16 different options: White, Black or African-American, Asian Indian, Chinese, Filipino, Japanese, Korean, Vietnamese, Other, Asian, Guamanian or Chamorro, Samoan, Other Pacific Islander, Some other race, and Hispanic. We factored and recoded these to White, Black, Hispanic, Asian, NHPI (Native Hawaiian or Pacific Islander), AIAN (American Indian, Alaska Native), and Other. We understand that these categories are problematic and contentious [@haney-lópez2006; @dowling2014], however they are the standard with quantitative data.

Similarly, gender was a contentious variable to account for. Although the GSS has several questions aimed at capturing respondent's gender, we found them to be lacking in breathe. For example, the GSS codes **sex** within a Male/Female binary based solely on interviewer discretion. Ultimately, we opted to use **sexnow1** since it was the variable that most closely captured self-described gender. However, it remains problematic as respondents were only provided four choice: Male, Female, Transgender, and None of these. Out of 2306 cases, only 10 were coded as Transgender and 9 as None of these.

Another important variable we sought to control for was citizenship status, since citizenship status and the associated stigma of illegality and threat of deportation have been found to increase fear in daily life [@golash-boza2015; @kline2019]. However, GSS data is not conducive to controlling for this variable due to the large number of NAs. Out of 2318 cases, 2164 were coded as NAs.

The second dimension we sought to capture was community-level characteristics, i.e., the social location of individuals. For this dimension, we drew on the following variables:

@@

# Data Script

@@

library(tidyverse)
library(dplyr)
library(tidyr)
library(here)
library(rio)
library(janitor)
library(knitr)
library(kableExtra)
library(mice)
library(forcats)
library(ggplot2)
library(RColorBrewer)
library(spelling)
```

@@

@@

@@

@@
